BayesPy: Variational Bayesian Inference in Python

### 2015-03-11 bayespy.tex
Modify to page limits

\jmlrheading{?}{2014}{?-?}{?/??}{?/??}{Jaakko Luttinen}

@@

\author{\name Jaakko Luttinen \email jaakko.luttinen@aalto.fi \\
       \addr Department of Information and Computer Science\\
       Aalto University, Finland}

@@

Bayesian framework provides theoretically solid and consistent way to construct
models and perform inference.  In practice, however, the inference is usually
analytically intractable and is therefore based on approximation methods such as
variational Bayes (VB), expectation propagation (EP) and Markov chain Monte
Carlo (MCMC) \citep{Bishop:2006}.  Deriving and implementing the formulas for an
approximation method is often straightforward but tedious, time consuming and
error prone.

BayesPy is a Python package providing tools for constructing Bayesian models and
performing variational Bayesian inference easily and efficiently.  It is based
on variational message passing (VMP) framework which defines a simple and local
message passing protocol \citep{Winn:2005}.  This enables implementation of small
general modules that can be used as building blocks for large complex models.
BayesPy offers a comprehensive collection of built-in blocks that can be used to
build a wide range of models and a simple interface for implementing custom
blocks.  The package is written for Python 3 and released under the GNU General
Public License v3.0 (GPLv3).

Several other projects have similar goals for making Bayesian inference easier
and faster to apply.  VB inference is available in Bayes Blocks
\citep{Raiko:2007}, VIBES \citep{Bishop:2002} and Infer.NET \citep{Infer.NET}.
Bayes Blocks is an open-source C++/Python package but it is limited to scalar
Gaussian nodes, a few deterministic functions and fully factorial posterior
approximation, thus making it very limited.  VIBES is an old software package
for Java, released under the revised BSD license, but it is no longer actively
developed.  VIBES has been replaced by Infer.NET, which supports a wide range of
posterior approximations.  However, Infer.NET is partly closed source and
licensed for non-commercial use only.  Instead of VB inference, mainly MCMC is
supported by other projects such as PyMC \citep{PyMC}, OpenBUGS \citep{OpenBUGS},
Dimple \citep{Dimple} and Stan \citep{Stan}.  Thus, there is a need for an open
source and maintained VB software package.

@@

BayesPy can be used to construct conjugate exponential family models.  The
documentation provides detailed examples of how to construct, for instance,
regression models, principal component analysis models, linear state-space
models, mixture models and hidden Markov models.  In addition, BayesPy has been
used in two publications about parameter expansion and time-varying dynamics for
linear state-space models \citep{Luttinen:2013,Luttinen:2014}, which show that
it is applicable to large complex models requiring efficient inference.

Using BayesPy for Bayesian inference consists of four main steps: constructing
the model, providing data, finding the posterior approximation and examining the
results.  The user constructs the model from small modular blocks called nodes.
Roughly, each node corresponds to a latent variable, a set of observations or a
deterministic function.  Some of the nodes may be set as observed by providing
data.  After that, inference engine is initialized and the message passing
algorithm is run in order to obtain the posterior approximation.  The resulting
posterior can be examined, for instance, by using a few built-in plotting
functions or printing the parameters of the posterior distributions.

Nodes are the key part of BayesPy as they are used to construct the model.
There are two types of nodes in BayesPy: stochastic and deterministic.
Stochastic nodes correspond to probability distributions and deterministic nodes
correspond to functions.  Built-in stochastic nodes include all common
exponential family distributions (e.g., Gaussian, gamma, Dirichlet and Poisson
distributions), a general mixture distribution and a few complex nodes for
dynamic variables (e.g., discrete and Gaussian Markov chains).  Built-in
deterministic nodes include a gating node and a general sum-product node.  In
case a model cannot be constructed using the built-in nodes, the documentation
provides instructions for implementing new nodes.

@@

This section demonstrates BayesPy with a Gaussian mixture model.  A simple
artificial dataset is created by drawing 200 samples from a 2-dimensional
Gaussian distribution.  80 samples have mean $[1, 1]$ and 120 samples have mean
$[0, 0]$:
\begin{lstlisting}
import numpy as np
N = 200, D = 2
data = np.random.randn(N, D)
data[:80,:] += np.ones(D)
\end{lstlisting}
We construct a mixture model for the data and assume that the parameters and the
cluster assignments are unknown, even the true number of clusters.  The model
uses a maximum number of five clusters but the effective number of clusters will
be determined automatically:
\begin{lstlisting}
K = 5
\end{lstlisting}
The unknown cluster means and precision matrices are given Gaussian and Wishart
prior distributions:
\begin{lstlisting}
from bayespy import nodes
mu = nodes.Gaussian(np.zeros(D), 0.01*np.identity(D), plates=(K,))
Lambda = nodes.Wishart(D, 0.01*np.identity(D), plates=(K,))
\end{lstlisting}
The \texttt{plates} keyword argument is used to define repetitions similarly as
the plate notation in graphical models.  The cluster assignments are categorical
variables, and the cluster probabilities are given a Dirichlet prior
distribution:
\begin{lstlisting}
alpha = nodes.Dirichlet(0.01*np.ones(K))
z = nodes.Categorical(alpha, plates=(N,))
\end{lstlisting}
The observations are from a Gaussian mixture distribution:
\begin{lstlisting}
y = nodes.Mixture(Z, nodes.Gaussian, mu, Lambda)
\end{lstlisting}
The second argument for the \texttt{Mixture} node defines the type of the
mixture distribution, in this case Gaussian.  The variable is marked as observed
by providing the data:
\begin{lstlisting}
y.observe(data)
\end{lstlisting}
Next, we want to find the posterior approximation for our latent variables.  We
create the variational Bayesian inference engine:
\begin{lstlisting}
from bayespy.inference import VB
Q = VB(y, mu, z, Lambda, alpha)
\end{lstlisting}
Before running the VMP algorithm, the symmetry in the model is broken by
a random initialization of the cluster assignments:
\begin{lstlisting}
z.initialize_from_random()
\end{lstlisting}
Without the random initialization, the clusters would not be separated.  The VMP
algorithm is run for 100 iterations or until convergence:
\begin{lstlisting}
Q.update(repeat=100)
\end{lstlisting}
The results can be examined, for instance, by printing the posterior
distributions of some variables:
\begin{lstlisting}
print(alpha)
print(mu)
\end{lstlisting}
In addition, module \texttt{bayespy.plot} contains several plotting functions to
examine the results visually.

@@

The results are summarized in Table~\ref{tab:speed}.  On the smaller datasets,
Infer.NET is slightly faster because BayesPy has some overhead from using
Python.  However, on the larger datasets, BayesPy is significantly faster
because it uses BLAS and LAPACK libraries for fast numerical
operations.\footnote{NumPy must use BLAS/LAPACK because BayesPy uses NumPy for
  numerical operations.}  In addition, if using multi-threaded BLAS/LAPACK, the
difference in wall-clock time can be even larger than in CPU time.  The speed of
BayesPy for these models could potentially be improved by using collapsed or
stochastic variational inference, and the accuracy could be improved by using
Gaussian-Wishart nodes which do not factorize with respect to the Gaussian and
Wishart variables.

@@

BayesPy provides a simple and efficient way to construct conjugate exponential
family models and to find the variational Bayesian posterior approximation in
Python.
%
Future plans for BayesPy include implementing more inference engines (e.g.,
maximum likelihood, expectation propagation and Gibbs sampling), improving the
VB engine (e.g., collapsed variational inference \citep{Hensman:2012} and
Riemannian conjugate gradient method \citep{Honkela:2010}) and implementing nodes
for non-parametric models (e.g., Gaussian processes and Dirichlet processes).

>>>>

\jmlrheading{?}{2015}{?-?}{?/??}{?/??}{Jaakko Luttinen}

@@

\author{\name Jaakko Luttinen \email jaakko.luttinen@aalto.fi \\
       \addr Department of Computer Science\\
       Aalto University, Finland}

@@

Bayesian framework provides a theoretically solid and consistent way to
construct models and perform inference.  In practice, however, the inference is
usually analytically intractable and is therefore based on approximation methods
such as variational Bayes (VB), expectation propagation (EP) and Markov chain
Monte Carlo (MCMC) \citep{Bishop:2006}.  Deriving and implementing the formulas
for an approximation method is often straightforward but tedious, time consuming
and error prone.

BayesPy is a Python 3 package providing tools for constructing models and
performing VB inference easily and efficiently.  It is based on the variational
message passing (VMP) framework which defines a simple message passing protocol
\citep{Winn:2005}.  This enables implementation of small general nodes that can
be used as building blocks for large complex models.  BayesPy offers a
comprehensive collection of built-in nodes that can be used to build a wide
range of models and a simple interface for implementing custom nodes.  The
package is released under the GNU General Public License v3.0 (GPLv3).

Several other projects have similar goals for making Bayesian inference easier
and faster to apply.  VB inference is available in Bayes Blocks
\citep{Raiko:2007}, VIBES \citep{Bishop:2002} and Infer.NET \citep{Infer.NET}.
Bayes Blocks is an open-source C++/Python package but limited to scalar Gaussian
nodes and a few deterministic functions, thus making it very limited.  VIBES is
an old software package for Java, released under the revised BSD license, but it
is no longer actively developed.  VIBES has been replaced by Infer.NET, which
%supports a wide range of posterior approximations.  However, Infer.NET
is partly closed source and licensed for non-commercial use only.  Instead of VB
inference, mainly MCMC is supported by other projects such as PyMC \citep{PyMC},
OpenBUGS \citep{OpenBUGS}, Dimple \citep{Dimple} and Stan \citep{Stan}.  Thus,
there is a need for an open source and maintained VB software package.

@@

BayesPy can be used to construct conjugate exponential family models.  The
documentation provides detailed examples of how to construct, for instance,
principal component analysis models, linear state-space models, mixture models
and hidden Markov models.  In addition, BayesPy has been used in two
publications about parameter expansion and time-varying dynamics for linear
state-space models \citep{Luttinen:2013,Luttinen:2014}.
% , which show that it is applicable to large complex models requiring efficient
% inference.

Using BayesPy for Bayesian inference consists of four main steps: constructing
the model, providing data, finding the posterior approximation and examining the
results.  The user constructs the model from small modular blocks called nodes.
Roughly, each node corresponds to a latent variable, a set of observations or a
deterministic function.
%Some of the nodes may be set as observed by providing
%data.
The inference engine is used to run the message passing algorithm in order to
obtain the posterior approximation.  The resulting posterior can be examined,
for instance, by using a few built-in plotting functions or printing the
parameters of the posterior distributions.

Nodes are the key part of BayesPy as they are used to construct the model.
There are two types of nodes in BayesPy: stochastic and deterministic.
Stochastic nodes correspond to probability distributions and deterministic nodes
correspond to functions.  Built-in stochastic nodes include all common
exponential family distributions (e.g., Gaussian, gamma and Dirichlet
distributions), a general mixture distribution and a few complex nodes for
dynamic variables (e.g., discrete and Gaussian Markov chains).  Built-in
deterministic nodes include a gating node and a general sum-product node.
%The
%documentation also provides instructions for implementing new nodes in case the
%built-in collection is insufficient.
In case a model cannot be constructed using the built-in nodes, the
documentation provides instructions for implementing new nodes.

@@

This section demonstrates BayesPy with a Gaussian mixture model.  A simple
artificial dataset is created by drawing 200 samples from a 2-dimensional
Gaussian distribution.  80 samples have mean $[1, 1]$ and 120 samples have mean
$[0, 0]$:
\begin{lstlisting}
import numpy as np
N = 200, D = 2
data = np.random.randn(N, D)
data[:80,:] += np.ones(D)
\end{lstlisting}
We construct a mixture model for the data and assume that the parameters, the
cluster assignments and the true number of clusters are unknown.  The model uses
a maximum number of five clusters but the effective number of clusters will be
determined automatically:
\begin{lstlisting}
K = 5
\end{lstlisting}
The unknown cluster means and precision matrices are given Gaussian and Wishart
prior distributions:
\begin{lstlisting}
from bayespy import nodes
mu = nodes.Gaussian(np.zeros(D), 0.01*np.identity(D), plates=(K,))
Lambda = nodes.Wishart(D, 0.01*np.identity(D), plates=(K,))
\end{lstlisting}
The \texttt{plates} keyword argument is used to define repetitions similarly as
the plate notation in graphical models.  The cluster assignments are categorical
variables, and the cluster probabilities are given a Dirichlet prior
distribution:
\begin{lstlisting}
alpha = nodes.Dirichlet(0.01*np.ones(K))
z = nodes.Categorical(alpha, plates=(N,))
\end{lstlisting}
The observations are from a Gaussian mixture distribution:
\begin{lstlisting}
y = nodes.Mixture(Z, nodes.Gaussian, mu, Lambda)
\end{lstlisting}
The second argument for the \texttt{Mixture} node defines the type of the
mixture distribution, in this case Gaussian.  The variable is marked as observed
by providing the data:
\begin{lstlisting}
y.observe(data)
\end{lstlisting}
Next, we want to find the posterior approximation for our latent variables.  We
create the variational Bayesian inference engine:
\begin{lstlisting}
from bayespy.inference import VB
Q = VB(y, mu, z, Lambda, alpha)
\end{lstlisting}
Before running the VMP algorithm, the symmetry in the model is broken by
a random initialization of the cluster assignments:
\begin{lstlisting}
z.initialize_from_random()
\end{lstlisting}
Without the random initialization, the clusters would not be separated.  The VMP
algorithm is run for 100 iterations or until convergence:
\begin{lstlisting}
Q.update(repeat=100)
\end{lstlisting}
The results can be examined, for instance, by printing the posterior
distributions of some variables:
\begin{lstlisting}
print(alpha)
print(mu)
\end{lstlisting}
In addition, module \texttt{bayespy.plot} contains several plotting functions to
examine the results visually.

@@

The results are summarized in Table~\ref{tab:speed}.  On the smaller datasets,
Infer.NET is slightly faster because BayesPy has some overhead of Python.
However, on the larger datasets, BayesPy is significantly faster because it uses
BLAS and LAPACK libraries for fast numerical operations.\footnote{Assuming that
  NumPy uses BLAS/LAPACK because BayesPy uses NumPy for numerical operations.}
In addition, if using multi-threaded BLAS and LAPACK, the difference in
wall-clock time can be even larger than in CPU time.  The speed of BayesPy for
these models could potentially be improved by using collapsed or stochastic
variational inference, and the accuracy could be improved by using
Gaussian-Wishart nodes which do not factorize with respect to the Gaussian and
Wishart variables.

@@

BayesPy provides a simple and efficient way to construct conjugate exponential
family models and to find the variational Bayesian posterior approximation in
Python.
%
In addition to the standard variational message passing, it supports several
advanced methods such as stochastic and collapsed variational inference.  Future
plans include support for non-conjugate models and non-parametric nodes (e.g.,
Gaussian and Dirichlet processes).
% Future plans for BayesPy include implementing more inference engines (e.g.,
% maximum likelihood, expectation propagation and Gibbs sampling), improving the
% VB engine (e.g., collapsed variational inference \citep{Hensman:2012} and
% Riemannian conjugate gradient method \citep{Honkela:2010}) and implementing nodes
% for non-parametric models (e.g., Gaussian processes and Dirichlet processes).
